r5r: rapid realistic routing on multimodal transport networks with R5 in R

### 2020-11-06 r-package/paper2020/paper2020.Rmd
General review of r5r paper.

<!-- Examples of similar papers -->
<!--  - https://transportfindings.org/article/8416-accessibility-toolbox-for-r-and-arcgis -->

@@

Transport routing is the process of finding the alternative routes and costs 
that connect places on a given transport network. Routing often involves complex
mathematical algorithms and it is a fundamental step required for transport
accessibility analysis, fleet allocation and transport simulation and planning
more broadly. Despite the long-standing importance of routing analysis to the 
field of transport and mobility studies, researchers and practitioners still 
frequently face practical challenges with this task. These challenges involve for
example the monetary costs of licensed software, data availability issues for data
hungry models, and the long computation times to run multiple routing scenarios, 
particularly in  large and complex multimodal transport networks.

While there is a growing number of open source routing models [@opentripplanneropentripplanner; 
@lovelace2019stplanr; @padgham2019dodgr], most options available do not efficiently 
process large transport networks. More importantly, these softwares generate
deterministic estimates of travel times without accounting for issues of service
reliability (REF) and they only return the fastest route option between a set of
origins and destinations. This simple approach overlooks how often people's 
actual route choices deviate from shortest travel time paths, what
can occur due to multiple factors such as imperfect information, monetary costs 
and safety issues [@larsen2011travel; @zhu2015people; @conway2019getting].

This paper presents [r5r](https://ipeagit.github.io/r5r/), a new open source 
R package for rapid realistic routing on multimodal transport networks based on
the [Rapid Realistic Routing on Real-world and Reimagined networks (R<sup>5</sup>)](https://github.com/conveyal/r5). 
R<sup>5</sup> is a powerful next-generation routing engine written in Java and
developed at Conveyal [@conway2017evidencebased; @conway2018accounting] to
replace its predecessor OpenTripPlanner. The r5r package provides a simple and
friendly interface to run R<sup>5</sup> locally from within R, what allows users
to efficiently calculate travel time matrices or generate multiple route 
alternatives between origins and destinations using seamless parallel computing.

@@

r5r has low data requirements, enabling routing analysis on different contexts 
wherever data on street networks from OpenStreetMap (OSM) is available. The 
package is also easily scalable allowing for fast computations at either the 
city or country level analysis. For multimodal transport routing, r5r will 
automatically combine OSM data with one or multiple public transport data sets 
provided by the user in  the standard GTFS format to create a routable transport
network.

@@

- `detailed_itineraries()`: computes detailed information on either the
shortest or multiple alternative routes between one or multiple origin/destination
pairs for a single departure time. The output includes detailed information on
route alternatives such as the transport mode, waiting time, travel time and 
distance of each segment of the trip. This function uses an R5-specific 
extension to the McRAPTOR routing algorithm to find optimal or less than optimal 
paths. The specific extension to McRAPTOR to do suboptimal path routing are not 
documented yet, but a detailed description of base McRAPTOR can be found in @delling2015roundbased.

Both routing functions are versatile so users can easily set customized inputs
such as transport modes, departure date and times, walking and cycling speeds, 
maximum trip duration, walking distances and number of public transport 
transfers, etc.

@@

The first step when using r5r is to set the amount of memory available to Java. 
By default `R` allocates only 512 MB of memory for Java processes. To increase 
available memory to 2GB, for example, we need to set the `java.parameters` 
option at the beginning of the script as follows:

@@

The package can be attached (with some other packages used in this article) with 
the following command:

@@

To illustrate functionality, the package includes a small sample data set for 
the city of Porto Alegre (Brazil). It includes three files:

- An OSM network in `.pbf` format.
- A public transport network in `GTFS.zip` format.
- The spatial coordinates of points covering the area in `.csv` format.
- The spatial coordinates of some of the city's points of interest.

@@

To build a routable transport network with r5r, the user only needs one command
line pointing to where the input data is stored:

```{r, message = FALSE}
r5r_core <- setup_r5(data_path)
```

@@

The presence of a `.pbf` file with OpenStreetMap data is mandatory because it is
required to create a street network. Meanwhile, a `GTFS.zip` file is only used to
calculate public transport trips, thus may not be necessary if that's not intended.
Multiple GTFS feeds can be used simultaneously, in which case R<sup>5</sup> will
merge all files into a single network automatically.

@@

Calculating a travel time matrix with r5r can be done very efficiently with the
`travel_time_matrix()` function. The function takes origins/destinations spatial
location (either as a spatial `sf POINT` object, or as a `data.frame` containing
the columns `id`, `lon` and `lat`) and travel parameters as inputs and outputs a
`data.table` with the travel time estimates between all combinations of origins 
and destinations that can be reached with the given parameters.

The `time_window` parameter is set to 1 by default, which means that travel time 
estimates will be calculated using a single departure time (specified by 
`departure_datetime`):

```{r, message = FALSE}
mode <- c("WALK", "TRANSIT")
max_walk_dist <- 1000
max_trip_duration <- 120
departure_datetime <- as.POSIXct("13-03-2019 14:00:00",
                                 format = "%d-%m-%Y %H:%M:%S")

# calculate travel time matrix
ttm <- travel_time_matrix(r5r_core,
                          origins = points,
                          destinations = points,
                          mode = mode,
                          departure_datetime = departure_datetime,
                          max_walk_dist = max_walk_dist,
                          max_trip_duration = max_trip_duration)

head(ttm)
```

Alternatively, a bigger time window can be specified. In such cases several 
travel time matrices are calculated using every minute from `departure_datetime` 
to the end of the given window as a distinct departure time. By default the 
function outputs the 50th percentile (i.e. the median) travel time between each 
origin and destination pair, which can be overruled by the `percentiles` 
parameter:

```{r, message = FALSE}
time_window <- 120
percentiles <- c(25, 50, 75)

# calculate travel time matrix
ttm <- travel_time_matrix(r5r_core,
                          origins = points,
                          destinations = points,
                          mode = mode,
                          departure_datetime = departure_datetime,
                          max_walk_dist = max_walk_dist,
                          max_trip_duration = max_trip_duration,
                          time_window = time_window,
                          percentiles = percentiles)

ttm
```

The image below depicts the travel time distribution for each origin/destination in which a trip could be completed within the specified parameters. The pairs are ordered in the x axis by their median travel time and each point in the plot area pictures a distinct distribution percentile.

```{r, message = FALSE}
percentiles <- 1:100

ttm <- travel_time_matrix(r5r_core,
                          origins = points,
                          destinations = points,
                          mode = mode,
                          departure_datetime = departure_datetime,
                          max_walk_dist = max_walk_dist,
                          max_trip_duration = max_trip_duration,
                          time_window = time_window,
                          percentiles = percentiles)

order_data <- ttm[, .(fromId, toId, travel_time_p025, travel_time_p050, travel_time_p075)]
order_data <- order_data[order(travel_time_p050, travel_time_p025, travel_time_p075)]
id_order   <- order_data[, id := paste0(fromId, "_", toId)]$id

@@

plot_data <- melt(ttm, measure = pctls, variable = "percentil", value = "travel_time")
plot_data[, percentil := as.integer(regmatches(percentil, regexpr("\\d+", percentil)))]
plot_data[, travel_time := fifelse(travel_time > max_trip_duration, NA_integer_, travel_time)]
plot_data[, id := factor(paste0(fromId, "_", toId), levels = id_order)]

ggplot(plot_data) + 
   geom_point(aes(id, travel_time), alpha = 0.02, size = 1) + 
   theme_minimal() +
   theme(axis.title.x = element_blank(), axis.text.x = element_blank())
```

@@

Often enough one wants more information on the itineraries between 
origin/destination pairs than simply the travel time. For such cases, the 
package includes the `detailed_itineraries()` function, which also takes travel 
restrictions as inputs, but outputs a much more detailed data table at the trip 
segment level. Another key feature of this function is that it (optionally) 
provides multiple alternative routes between origin/destination pairs:

@@

# set restrictions
mode <- c("WALK", "TRANSIT")
max_walk_dist <- Inf
max_trip_duration <- 120
departure_datetime <- as.POSIXct("13-03-2019 14:00:00",
                                 format = "%d-%m-%Y %H:%M:%S")

@@

dit
```

Since the `shortest_path` argument has been set to `FALSE`, `dit` holds 
information for multiple itineraries between the same origin/destination pairs:

@@

### Visualize results

@@

The [R<sup>5</sup> routing engine](https://github.com/conveyal/r5) is developed 
at [Conveyal](https://www.conveyal.com/) with contributions from several developers. 
This work was supported by the Brazilian Institute for Applied Economic Research 
(Ipea).

@@

<!-- Rafa's notes -->
<!-- jean: #>-  mass of assumptions regarding decision-making and routing (Miller, 2018).  -->
<!-- Haugen, 2011; Schwanen, 2008 -->

@@

<!-- This article presents **r5r**, a new open-source `R` package for rapid  -->
<!-- realistic routing on multimodal transport networks. -->

>>>>

<!-- Examples of similar papers -->

<!--  - https://transportfindings.org/article/8416-accessibility-toolbox-for-r-and-arcgis -->

@@

Transport routing is the process of finding alternative routes and costs that connect places on a given transport network. Routing often involves complex mathematical algorithms and it is a fundamental step required for transport accessibility analysis, fleet allocation and transport simulation and planning more broadly. Despite the long-standing importance of routing analysis to the field of transport and mobility studies, researchers and practitioners still frequently face practical challenges when carrying out this task. These challenges involve, for example, the monetary costs of licensed software, limited data availability for data hungry models, and the long computation times required to run multiple routing scenarios, particularly in large and complex multimodal transport networks.

While there is a growing number of open source routing models [@opentripplanneropentripplanner; @lovelace2019stplanr; @padgham2019dodgr], most options available do not efficiently process large transport networks. More importantly, these softwares generate deterministic estimates of travel times without accounting for variability in results (due to service reliability issues, for example) and only return the fastest route option between a set of origins and destinations (see [@conway2018accounting] for a more in-depth discussion). This simple approach overlooks how often people's actual route choices deviate from shortest travel time paths, what can occur due to multiple factors such as imperfect information, monetary costs and safety issues [@larsen2011travel; @zhu2015people; @conway2019getting].

This paper presents [r5r](https://ipeagit.github.io/r5r/), a new open source R package for rapid realistic routing on multimodal transport networks based on the [Rapid Realistic Routing on Real-world and Reimagined networks (R](https://github.com/conveyal/r5)<sup>5</sup>). R<sup>5</sup> is a powerful next-generation routing engine written in Java and developed at Conveyal [@conway2017evidencebased; @conway2018accounting] to replace its predecessor OpenTripPlanner. The r5r package provides a simple and friendly interface to run R<sup>5</sup> locally from within R, which allows users to efficiently calculate travel time matrices or generate multiple route alternatives between origins and destinations using seamless parallel computing.

@@

The r5r package has low data requirements, enabling routing analysis on different contexts wherever data on street networks from [OpenStreetMap](https://www.openstreetmap.org/) (OSM) is available. The package is also easily scalable allowing for fast computation at either the city or country levels of analysis. For multimodal transport routing, r5r will automatically combine OSM data with one or multiple public transport data sets provided by the user in the standard [General Transit Feed Specification](https://developers.google.com/transit/gtfs/) (GTFS) format to create a routable transport network.

@@

-   `detailed_itineraries()`: computes detailed information on either the shortest or multiple alternative routes between one or multiple origin/destination pairs for a single departure time. The output includes detailed information on route alternatives such as the transport mode, waiting time, travel time and distance of each segment of the trip. This function uses an R5-specific extension to the McRAPTOR routing algorithm to find optimal or less than optimal paths. The specific extension to McRAPTOR to do suboptimal path routing are not documented yet, but a detailed description of base McRAPTOR can be found in [@delling2015roundbased].

Both routing functions are versatile so users can easily set customized inputs such as transport modes, departure date and times, walking and cycling speeds, maximum trip duration, walking distances and number of public transport transfers.

@@

The first step when using r5r is to set the amount of memory available to Java. By default `R` allocates only 512 MB of memory to the Java Virtual Machine (JVM), which is not enough for the tasks r5r is designed to carry out, such as calculating travel time matrices on large multimodal networks. To increase available memory to 2GB[^1], for example, the user needs to set the `java.parameters` option at the beginning of the script as follows:

[^1]: Note that large study areas may require significantly more memory to run properly.

@@

The package can be attached (alogside some other packages used in this article) with the following command:

@@

To illustrate its functionalities, the package includes sample datasets for the cities of São Paulo and Porto Alegre (both in Brazil). Each dataset includes three files:

-   An OSM network in `.pbf` format.
-   A public transport network in `GTFS.zip` format.
-   The spatial coordinates of points covering the area in `.csv` format.

Additionally, the Porto Alegre dataset includes a `.csv` file containing the spatial coordinates of some of the city's points of interest. A sample of the Porto Alegre dataset, which will be used in the examples of this paper, is presented below:

@@

To build a routable transport network with r5r and load it in memory, the user needs one command line pointing to the location on disk where the input data is stored:

```{r, message = FALSE}
r5r_core <- setup_r5(data_path, verbose = FALSE)
```

@@

The `.pbf` file with OpenStreetMap data is mandatory because it is required to create a street network for the study area. The `GTFS.zip` file, conversely, is only required for calculating public transport trips, thus it is not necessary if that's not intended. Multiple GTFS feeds can be used simultaneously, in which case R<sup>5</sup> will merge all files into a single public transport network automatically.

@@

Calculating a travel time matrix with r5r can be done with the `travel_time_matrix()` function. This function takes, as inputs, origins/destinations spatial location (either as a spatial `sf POINT` object, or as a `data.frame` containing the columns `id`, `lon` and `lat`) and travel parameters and outputs a `data.table` with the travel time estimates between all combinations of origins and destinations that can be reached with the given parameters. An example of the `travel_time_matrix()` function's usage is presented as follows:

```{r, message = FALSE}
mode <- c("WALK", "TRANSIT")
max_walk_dist <- 1000
max_trip_duration <- 120
departure_datetime <- as.POSIXct("13-05-2019 14:00:00",
                                 format = "%d-%m-%Y %H:%M:%S")
# calculate travel time matrix
ttm <- travel_time_matrix(r5r_core,
                          origins = points,
                          destinations = points,
                          mode = mode,
                          departure_datetime = departure_datetime,
                          max_walk_dist = max_walk_dist,
                          max_trip_duration = max_trip_duration,
                          verbose = FALSE)

ttm
```
The `time_window` parameter is set to 1 by default, which means that travel time estimates are calculated using a single departure time (specified by the `departure_datetime` parameter). Alternatively, a bigger time window can be specified. In such cases several travel time matrices are calculated using every minute from `departure_datetime` to the end of the given window as a distinct departure time. By default the function outputs the 50th percentile (i.e. the median) travel time between each origin and destination pair, which can be overruled by the `percentiles` parameter.

```{r, message = FALSE}
mode <- c("WALK", "TRANSIT")
max_walk_dist <- 1000
max_trip_duration <- 120
departure_datetime <- as.POSIXct("13-05-2019 14:00:00",
                                 format = "%d-%m-%Y %H:%M:%S")
time_window <- 120
percentiles <- 1:100

# calculate travel time matrix
ttm <- travel_time_matrix(r5r_core,
                          origins = points,
                          destinations = points,
                          mode = mode,
                          departure_datetime = departure_datetime,
                          max_walk_dist = max_walk_dist,
                          max_trip_duration = max_trip_duration,
                          time_window = time_window,
                          percentiles = percentiles,
                          verbose = FALSE)

ttm[, .(fromId, toId, travel_time_p025, travel_time_p050, travel_time_p075)]
```

The plot below depicts the travel time distribution for each origin/destination in which a trip could be completed within the specified parameters. The pairs are ordered in the x axis by their median travel time and each point in the plot area represents a distinct distribution percentile.

```{r, message = FALSE}
order_data <- ttm[, .(fromId, toId, travel_time_p025, travel_time_p050, travel_time_p075)]
order_data <- order_data[order(travel_time_p050, travel_time_p025, travel_time_p075)]
id_order   <- order_data[, id := paste0(fromId, "_", toId)]$id

@@

plot_data <- melt(ttm, measure = pctls, variable = "percentil", value = "travel_time", 
                  na.rm = TRUE)
plot_data[, percentil := as.integer(regmatches(percentil, regexpr("\\d+", percentil)))]
plot_data[, id := factor(paste0(fromId, "_", toId), levels = id_order)]

ggplot(plot_data) + 
   geom_point(aes(id, travel_time), alpha = 0.005, size = 0.5) +
   expand_limits(y = 120) +
   scale_y_continuous(breaks = c(0, 30, 60, 90, 120)) +
   theme(axis.title.x = element_blank(), 
         axis.text.x = element_blank(),
         axis.ticks = element_blank(),
         panel.border = element_rect(fill = NA, colour = "grey80", size=1)) +
   labs(title = "Travel time variation among percentiles, per OD pair",
        y = "Travel Time (min)")
```

@@

Often enough one wants more information on the itineraries between origin/destination pairs than simply the travel time. For such cases, r5r includes the `detailed_itineraries()` function. This function also takes travel restrictions as inputs, but outputs a more detailed table of trips disaggregated by segment. Similarly to the `travel_time_matrix()` function, the `origins` and `destinations` parameters of `detailed_itineraries()` can be either spatial `sf POINT` objects, or `data.frames` containing the columns `id`, `lon` and `lat`. Itineraries are calculated between matching pairs of locations in both datasets, so `origins` and `destinations` inputs must contain the same number of observations. However, if the input datasets do not match but `origins` contains only one observation, it is automatically expanded to match the number of observations in `destinations`. The calculations are automatically made in parallel, thus optimizing the use of the available computational resources.

@@

# set restrictions
mode <- c("WALK", "TRANSIT")
max_walk_dist <- Inf
max_trip_duration <- 120
departure_datetime <- as.POSIXct("13-05-2019 14:00:00",
                                 format = "%d-%m-%Y %H:%M:%S")

@@

st_set_geometry(dit, NULL)
```
A key feature of `detailed_itineraries()` is that it (optionally) provides multiple alternative routes between origin/destination pairs, when the `shortest_path` parameter is set to `FALSE`. This is the case in the example above, so `dit` holds information for multiple itineraries between the same origin/destination pairs, as can be seen below.

@@

#### Visualize results

@@

The [R](https://github.com/conveyal/r5)<sup>5</sup> routing engine is developed at [Conveyal](https://www.conveyal.com/) with contributions from several developers. This work was supported by the Brazilian Institute for Applied Economic Research (Ipea).

@@

<!-- Rafa's notes -->

<!-- jean: #>-  mass of assumptions regarding decision-making and routing (Miller, 2018).  -->

<!-- Haugen, 2011; Schwanen, 2008 -->

@@

<!-- This article presents **r5r**, a new open-source `R` package for rapid  -->

<!-- realistic routing on multimodal transport networks. -->
